---
title: "Understanding community assemblage using species and functional diversity across elevational gradients from the tropics to the arctic"
format:
    html:
      self-contained: true
  # pdf:
  #   documentclass: article
editor: visual
execute:
  echo: false
  error: true
  out-width: '100%'
---

```{r}
#| label: set-up
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(targets)
  library(tarchetypes)
  library(gt)
  library(dplyr)
  library(tidyr)
  library(patchwork)
})

```

## Introduction

Species and functional diversity are different metrics describing biodiversity and ecosystem function and they are affected by both abiotic and biotic filtering as well as the regional species pool. These metrics can be used to understand how plant communities respond to different filters and how communities assemble along broad environmental gradients.

We assessed plant Shannon diversity and functional traits (i.e. community weighted means and variance in leaf functional traits) along five elevational gradients in the arctic (Svalbard), boreal (Norway), temperate (Colorado), sub-tropics (China) and tropic mountains (Peru). Specifically, we ask how plant communities respond to the environmental drivers along elevational gradients in terms of species and functional diversity, and how consistent these patterns are across a broad latitudinal gradient from the tropics to the arctic. And what is the importance of intraspecific trait variation.

## Methods

### Study systems and sites

Description of the study systems and sites.

We studied the following ecosystems/countries: arctic (PFTC4 Svalbard), boreal (PFTC6 Norway), temperate (Colorado), sub-tropics (PFTC1 and 2 China) and tropics (PFTC3, Puna project and PFCT5 Peru). 

```{r}
#| label: fig-study-sites-map
#| echo: false
#| warning: false
#| message: false
#| fig-height: 6
#| fig-cap: "World map showing all the regions."

tar_load(regions_world_map)
regions_world_map

```

### Data collection

We collected data on species community composition and functional leaf trait data from all PFTC courses and RMBL in Colorado. Describe data collection and refer to data papers.
Describe which gradinets were used, e.g. Svalbard only control, not bird cliff, Peru not recently burned, etc.
There are no chemical traits from Norway and South Africa yet, and these countries were excluded from the multivariate analysis.

Climate data: site-level mean latitude (plot coordinates averaged within each site) versus downscaled mean annual temperature at 2 m (T2m), using the same climate extract as in the analyses.


### Data analysis

Shannon diversity was calcualted using the vegan pacakge (REF).

Size traits were log transformed. Trait means and higher moments were calculated using a trait imputation and bootstrapping with traitstrap (Maitner et al. 2023).

PCA

Run mixed-effects models for diversity and traits vs. climate and latitude. Run a global model across all regions and regional models for each region.

```{r}
#| label: fig-downscaled-t2m-latitude
#| echo: false
#| warning: false
#| message: false
#| fig-height: 4
#| fig-cap: "Downscaled mean annual temperature versus latitude. The colours show the regions and point size shows Elevation percentile, with low values indicate low elevation."

tar_load(downscaled_t2m_latitude_fig)
downscaled_t2m_latitude_fig

```

\newpage

## Results

### Plant diversity

Shannon diversity decreases with latitude and shows no significant relationship with mean annual temperature or vapour pressure deficit.

```{r}
#| label: fig-diversity-panels
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 11
#| fig-cap: "Shannon diversity versus latitude, mean annual temperature and vapour pressure deficit (VPD). The points show the raw data and the lines are predictions. Solid lines indicare significant relationships, while dashed lines are non-significnat. The colours indicate the regions."

tar_load(diversity_three_panel_fig)
diversity_three_panel_fig

```

### Functional leaf traits

- Size traits decrease with elevation

- LDMC increases and SLA decreases with elevation

- Carbon and nutrient show different patterns with elevation across the regions.


```{r}
#| label: fig-trait-distribution
#| echo: false
#| warning: false
#| message: false
#| fig-height: 12
#| fig-width: 8
#| fig-cap: "Distribution of trait values for different elevational bands. The colours indicated the regions. The size traits are log transformed."

tar_load(trait_distribution_ridgeline_fig)
trait_distribution_ridgeline_fig

```

The PCA for all size and leaf economic traits show that these traits are perpendicular to each other. The regions separate well across the leaf economic trait axis, with colder sites having higher SLA and lower LDMC (aquisitative traits). Warmer sites are generally taller and have more conservative traits.

```{r}
#| label: fig-trait-pca-full
#| echo: false
#| warning: false
#| fig-height: 6
#| fig-cap: "Principle component analysis for size and leaf economic traits. The regions are coloured and an elipsoid is drawn for each region. The arrows indicate the loadings for each trait. The linetype indicates size (solid) and leaf economic traits (dashed). The point size indiciates the elevational percentile within each region."

tar_load(trait_pca_full_fig)
trait_pca_full_fig

```

\newpage

The patterns are more clear when including chemical traits and itsotopes. PC axis 1 correlates with leaf economic traits (SLA, LDMC, leaf C, N, P, and leaf thickness), while PC axis 2 correlates with size traits and dN15 isotoeps. Colder sites have smaller leaves and higher SLA (= faster traits). Warmer sites have larger leaves and higher LDMC (= slower traits).

```{r}
#| label: fig-trait-pca
#| echo: false
#| warning: false
#| fig-height: 6
#| fig-cap: "Principle component analysis for size, leaf economic, and isotope traits. The regions are coloured and an elipsoid is drawn for each region. The arrows indicate the loadings for each trait. The linetype indicates size (solid), leaf economic (dashed), and isotope traits (dotted). The point size indiciates the elevational percentile within each region."

tar_load(trait_pca_fig)
trait_pca_fig

```

\newpage

- Plant height, dry mass, leaf area and dC13 show a unimodal global relationship with mean annual temperature, showing highest values at intermediate temperatures

- NP decreases and dN15 increases globally with mean annual temperature

- The regional models show a close correlation with the global model for dry mass and leaf area.

- CN shows significant regaional relationship with mean annual temperature, but showing a reversed unimodal relationship (U shaped)

- Leaf P decreases with mean annual temperature, but only regionally

- Leaf thickness, SLA, LDMC, leaf C, and N show no global or regional relationship with mean annual temperature

```{r}
#| label: fig-trait-ds-t2m
#| echo: false
#| warning: false
#| fig-height: 9
#| fig-width: 12
#| fig-cap: "Functional traits versus mean annual tempearture. The points show the raw data and the lines are predictions. Solid lines indicare significant relationships, while dashed lines are non-significnat. The colours indicate the regions."

tar_load(trait_climate_ds_t2m_fig)
trait_climate_ds_t2m_fig

```

\newpage

- Leaf thickness decreases globally with VPD. SLA, Leaf N, and P show a unimodal global relationship with VPD, while LDMC, CN, NP (highest values at intermediate VPD), and dC13 show a reversed unimodal relationship (U shaped; lowest values at intermediate VPD).

- Dry mass, leaf area and SLA show regional unimodal relationships with VPD.

- Leaf thickness and CN ratio show a reveresed regional unimodal relationship with VPD.

- Leaf P decreases regionally with VPD, showing opposite pattern to the global relationship.

- The regional pattern for leaf thickness, SLA, and CN ratio sort of reflects the global pattern.

- Plant height, leaf C and dN15 show no global or regional relationship with VPD.

```{r}
#| label: fig-trait-ds-vpd
#| echo: false
#| warning: false
#| fig-height: 9
#| fig-width: 12
#| fig-cap: "Functional traits versus vapour pressure deficit. The points show the raw data and the lines are predictions. Solid lines indicare significant relationships, while dashed lines are non-significnat. The colours indicate the regions."

tar_load(trait_climate_ds_vpd_fig)
trait_climate_ds_vpd_fig

```

### Intraspecific trait variation (ITV)

Trait sampling coverage by region. We do not have plot level trait data from China, and partly from Colorado and Norway. Comparing ITV between regions is therefore complicated.

```{r}
#| label: fig-trait-coverage
#| echo: false
#| warning: false
#| fig-height: 6
#| fig-cap: "Proportion of trait data sampled at the plot, site, gradient and regional level shown by region."

tar_load(trait_coverage_fig)
trait_coverage_fig

```

```{r}
#| label: fig-itv-variance
#| echo: false
#| warning: false
#| fig-height: 6
#| fig-cap: "Variance partitioning for different plant functional traits, split into between region, sites, species, and within species. Note that this analysis is done on raw trait values and not on CWM."

tar_load(itv_variance_plot)
itv_variance_plot

```

```{r}
#| label: fig-leps-itv
#| echo: false
#| warning: false
#| fig-height: 6
#| fig-width: 12

tar_load(leps_itv_plot)
leps_itv_plot

```

## Trait Model Results

```{r}
#| label: tab-trait-results
#| echo: false
#| warning: false
#| message: false

tar_load(trait_models_output)

# Create a nice table of trait model results from the new long-format models
trait_results_table <- trait_models_output |>
    # Unnest the tidy results
    unnest(tidy_results) |>
    filter(effect == "fixed", term != "(Intercept)") |> # Only fixed effects, exclude intercept
    mutate(
        # Clean up trait names using fancy names
        trait_clean = case_when(
            trait_trans == "plant_height_cm_log" ~ "Height cm",
            trait_trans == "dry_mass_g_log" ~ "Dry mass g",
            trait_trans == "leaf_area_cm2_log" ~ "Area cm²",
            trait_trans == "thickness_mm_log" ~ "Thickness mm",
            trait_trans == "ldmc" ~ "LDMC",
            trait_trans == "sla_cm2_g" ~ "SLA cm²/g",
            trait_trans == "c_percent" ~ "C %",
            trait_trans == "n_percent" ~ "N %",
            TRUE ~ stringr::str_replace_all(trait_trans, "_", " ") |> stringr::str_to_title()
        ),
        # Clean up predictor names using the climate_variable_clean from the data
        predictor_clean = case_when(
            climate_variable == "ds_t2m" ~ "Mean annual temperature at 2 m (downscaled)",
            climate_variable == "ds_vpd" ~ "Vapour pressure deficit (downscaled)",
            climate_variable == "growing_season_length" ~ "Growing Season Length (GEE)",
            climate_variable == "gsl_1981-2010_chelsa" ~ "Growing Season Length (CHELSA)",
            climate_variable == "gst_1981-2010_chelsa" ~ "Growing Season Temperature (CHELSA)",
            climate_variable == "gsp_1981-2010_chelsa" ~ "Growing Season Precipitation (CHELSA)",
            climate_variable == "pet_penman_mean_1981-2010_chelsa" ~ "Potential Evapotranspiration (CHELSA)",
            climate_variable == "vpd_mean_1981-2010_chelsa" ~ "Vapour Pressure Deficit (CHELSA)",
            climate_variable == "mean_temperture_warmest_quarter_bioclim" ~ "Mean Temperature Warmest Quarter (WorldClim)",
            climate_variable == "precipitation_warmest_quarter_bioclim" ~ "Precipitation Warmest Quarter (WorldClim)",
            climate_variable == "diurnal_range_bioclim" ~ "Mean Diurnal Range (WorldClim)",
            climate_variable == "annual_temperature_bioclim" ~ "Annual Temperature (WorldClim)",
            TRUE ~ climate_variable
        ),
        # Round numeric values
        estimate = round(estimate, 3),
        std.error = round(std.error, 3),
        statistic = round(statistic, 2),
        p.value = round(p.value, 3),
        # Create significance indicator for formatting
        is_significant = p.value < 0.05
    ) |>
    select(trait_clean, predictor_clean, estimate, std.error, statistic, df, p.value, is_significant)

# Create the gt table with proper ordering
trait_results_table |>
    arrange(predictor_clean, trait_clean) |> # Sort by predictor first, then trait
    group_by(predictor_clean) |>
    gt() |>
    cols_hide(columns = c(predictor_clean, is_significant)) |> # Hide predictor and significance columns
    tab_header(
        title = "Trait Model Results",
        subtitle = "Mixed-effects models showing relationships between functional traits and environmental predictors"
    ) |>
    cols_label(
        trait_clean = "Trait",
        estimate = "Estimate",
        std.error = "SE",
        statistic = "t-statistic",
        df = "df",
        p.value = "p-value"
    ) |>
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_body(
            columns = everything(),
            rows = is_significant == TRUE
        )
    ) |>
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_column_labels()
    ) |>
    tab_style(
        style = cell_text(align = "center"),
        locations = cells_body(columns = c(estimate, std.error, statistic, df, p.value))
    ) |>
    tab_options(
        table.font.size = 11,
        heading.title.font.size = 14,
        heading.subtitle.font.size = 12,
        column_labels.font.weight = "bold"
    )

```