---
title: "Understanding community assemblage using species and functional diversity across elevational gradients from the tropics to the arctic"
format:
    html:
      self-contained: true
  # pdf:
  #   documentclass: article
editor: visual
execute:
  echo: false
  error: true
  out-width: '100%'
---

```{r}
#| label: set-up
#| echo: false
#| message: false
#| warning: false

suppressPackageStartupMessages({
  library(targets)
  library(tarchetypes)
  library(gt)
  library(dplyr)
  library(tidyr)
  library(patchwork)
})

```

## Introduction

Species and functional diversity are different metrics describing biodiversity and ecosystem function and they are affected by both abiotic and biotic filtering as well as the regional species pool. These metrics can be used to understand how plant communities respond to different filters and how communities assemble along broad environmental gradients.

We assessed plant Shannon diversity and functional traits (i.e. community weighted means and variance in leaf functional traits) along five elevational gradients in the arctic (Svalbard), boreal (Norway), temperate (Colorado), sub-tropics (China) and tropic mountains (Peru). Specifically, we ask how plant communities respond to the environmental drivers along elevational gradients in terms of species and functional diversity, and how consistent these patterns are across a broad latitudinal gradient from the tropics to the arctic. And what is the importance of intraspecific trait variation.

## Methods

### Study systems and sites

Description of the study systems and sites.

We studied the following ecosystems/countries: arctic (PFTC4 Svalbard), boreal (PFTC6 Norway), temperate (Colorado), sub-tropics (PFTC1 and 2 China) and tropics (PFTC3, Puna project and PFCT5 Peru). 

```{r}
#| label: fig-study-sites-map
#| echo: false
#| warnings: false
#| message: false
#| fig-height: 6
#| fig-cap: "World map showing all the regions."

tar_load(regions_world_map)
regions_world_map

```

### Data collection

We collected data on species community composition and functional leaf trait data from all PFTC courses and RMBL in Colorado. Describe data collection and refer to data papers.
Describe which gradinets were used, e.g. Svalbard only control, not bird cliff, Peru not recently burned, etc.
There are no chemical traits from Norway and South Africa yet, and these countries were excluded from the multivariate analysis.

Climate data: site-level mean latitude (plot coordinates averaged within each site) versus downscaled mean annual temperature at 2 m (T2m), using the same climate extract as in the analyses.


### Data analysis

Shannon diversity was calcualted using the vegan pacakge (REF).

Size traits were log transformed. Trait means and higher moments were calculated using a trait imputation and bootstrapping with traitstrap (Maitner et al. 2023).

PCA

Run mixed-effects models for diversity and traits vs. climate and latitude. Run a global model across all regions and regional models for each region.

```{r}
#| label: fig-downscaled-t2m-latitude
#| echo: false
#| warning: false
#| message: false
#| fig-height: 4
#| fig-cap: "Downscaled mean annual temperature versus latitude. The colours show the regions and point size shows Elevation percentile, with low values indicate low elevation."

tar_load(downscaled_t2m_latitude_fig)
downscaled_t2m_latitude_fig

```

\newpage

## Results

### Plant diversity

Three-panel summary of Shannon diversity (cover-based): latitude (scaled predictor), downscaled mean annual temperature at 2 m (T2m), and vapour pressure deficit (VPD). Points are observations; grey lines and ribbons are mixed-model marginal predictions. Regions use colours; linetype encodes significance of the predictor slope.

```{r}
#| label: fig-diversity-panels
#| echo: false
#| warnings: false
#| fig-height: 5
#| fig-width: 11

tar_load(diversity_three_panel_fig)
diversity_three_panel_fig

```

### Functional leaf traits

This ridgeline plot shows the distribution of size, leaf economic and isotope traits along the elevation gradient.

```{r}
#| label: fig-trait-distribution
#| echo: false
#| warnings: false
#| message: false
#| fig-height: 12
#| fig-width: 8
#| fig-cap: "Distribution of log transformed trait values for different elevational bands. The colours indicated the regions."

tar_load(trait_distribution_ridgeline_fig)
trait_distribution_ridgeline_fig

```

This figure shows the multivariate trait space for a subset of traits available across all regions. The ecosystems are well separated in trait space, with clear differentiation between regions.

```{r}
#| label: fig-trait-pca-full
#| echo: false
#| warnings: false
#| fig-height: 6
#| fig-cap: "Principle component analysis for size and leaf economic traits. The regions are coloured and an elipsoid shows is drawn for each region. The arrows indicate the loadings for each trait. The linetype indicates size (solid) and leaf economic traits (dashed). The point size indiciates the elevational percentile within each region."

tar_load(trait_pca_full_fig)
trait_pca_full_fig

```

\newpage

This figure shows the multivariate trait space including chemical traits. Norway is not part of this figure, because we do not have the chemical traits yet. The ecosystems are well separated in trait space, with the temperate and sub-tropic regions showing some overlap.

```{r}
#| label: fig-trait-pca
#| echo: false
#| warnings: false
#| fig-height: 6
#| fig-cap: "Principle component analysis for size, leaf economic, and isotope traits. The regions are coloured and an elipsoid shows is drawn for each region. The arrows indicate the loadings for each trait. The linetype indicates size (solid), leaf economic (dashed), and isotope traits (dotted)."

tar_load(trait_pca_fig)
trait_pca_fig

```

\newpage

These figures show community-weighted mean functional traits (all analysed traits, mean bootstrap estimates only) versus **downscaled** climate at site resolution: mean annual temperature at 2 m (T2m), then vapour pressure deficit (VPD). Grey lines are global marginal predictions from mixed-effects models; coloured lines and ribbons are regional predictions (same plotting code for both panels).

```{r}
#| label: fig-trait-ds-t2m
#| echo: false
#| warnings: false
#| fig-height: 9
#| fig-width: 12

tar_load(trait_climate_ds_t2m_fig)
trait_climate_ds_t2m_fig

```

\newpage

```{r}
#| label: fig-trait-ds-vpd
#| echo: false
#| warnings: false
#| fig-height: 9
#| fig-width: 12

tar_load(trait_climate_ds_vpd_fig)
trait_climate_ds_vpd_fig

```

### The importance of intraspecific trait variation (ITV)

```{r}
#| label: fig-itv-variance
#| echo: false
#| warnings: false
#| fig-height: 6

tar_load(itv_variance_plot)
itv_variance_plot

```

```{r}
#| label: fig-leps-itv
#| echo: false
#| warnings: false
#| fig-height: 6
#| fig-width: 12

tar_load(leps_itv_plot)
leps_itv_plot

```

## Trait Model Results

```{r}
#| label: tab-trait-results
#| echo: false
#| warning: false
#| message: false

tar_load(trait_models_output)

# Create a nice table of trait model results from the new long-format models
trait_results_table <- trait_models_output |>
    # Unnest the tidy results
    unnest(tidy_results) |>
    filter(effect == "fixed", term != "(Intercept)") |> # Only fixed effects, exclude intercept
    mutate(
        # Clean up trait names using fancy names
        trait_clean = case_when(
            trait_trans == "plant_height_cm_log" ~ "Height cm",
            trait_trans == "dry_mass_g_log" ~ "Dry mass g",
            trait_trans == "leaf_area_cm2_log" ~ "Area cm²",
            trait_trans == "thickness_mm_log" ~ "Thickness mm",
            trait_trans == "ldmc" ~ "LDMC",
            trait_trans == "sla_cm2_g" ~ "SLA cm²/g",
            trait_trans == "c_percent" ~ "C %",
            trait_trans == "n_percent" ~ "N %",
            TRUE ~ stringr::str_replace_all(trait_trans, "_", " ") |> stringr::str_to_title()
        ),
        # Clean up predictor names using the climate_variable_clean from the data
        predictor_clean = case_when(
            climate_variable == "ds_t2m" ~ "Mean annual temperature at 2 m (downscaled)",
            climate_variable == "ds_vpd" ~ "Vapour pressure deficit (downscaled)",
            climate_variable == "growing_season_length" ~ "Growing Season Length (GEE)",
            climate_variable == "gsl_1981-2010_chelsa" ~ "Growing Season Length (CHELSA)",
            climate_variable == "gst_1981-2010_chelsa" ~ "Growing Season Temperature (CHELSA)",
            climate_variable == "gsp_1981-2010_chelsa" ~ "Growing Season Precipitation (CHELSA)",
            climate_variable == "pet_penman_mean_1981-2010_chelsa" ~ "Potential Evapotranspiration (CHELSA)",
            climate_variable == "vpd_mean_1981-2010_chelsa" ~ "Vapour Pressure Deficit (CHELSA)",
            climate_variable == "mean_temperture_warmest_quarter_bioclim" ~ "Mean Temperature Warmest Quarter (WorldClim)",
            climate_variable == "precipitation_warmest_quarter_bioclim" ~ "Precipitation Warmest Quarter (WorldClim)",
            climate_variable == "diurnal_range_bioclim" ~ "Mean Diurnal Range (WorldClim)",
            climate_variable == "annual_temperature_bioclim" ~ "Annual Temperature (WorldClim)",
            TRUE ~ climate_variable
        ),
        # Round numeric values
        estimate = round(estimate, 3),
        std.error = round(std.error, 3),
        statistic = round(statistic, 2),
        p.value = round(p.value, 3),
        # Create significance indicator for formatting
        is_significant = p.value < 0.05
    ) |>
    select(trait_clean, predictor_clean, estimate, std.error, statistic, df, p.value, is_significant)

# Create the gt table with proper ordering
trait_results_table |>
    arrange(predictor_clean, trait_clean) |> # Sort by predictor first, then trait
    group_by(predictor_clean) |>
    gt() |>
    cols_hide(columns = c(predictor_clean, is_significant)) |> # Hide predictor and significance columns
    tab_header(
        title = "Trait Model Results",
        subtitle = "Mixed-effects models showing relationships between functional traits and environmental predictors"
    ) |>
    cols_label(
        trait_clean = "Trait",
        estimate = "Estimate",
        std.error = "SE",
        statistic = "t-statistic",
        df = "df",
        p.value = "p-value"
    ) |>
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_body(
            columns = everything(),
            rows = is_significant == TRUE
        )
    ) |>
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_column_labels()
    ) |>
    tab_style(
        style = cell_text(align = "center"),
        locations = cells_body(columns = c(estimate, std.error, statistic, df, p.value))
    ) |>
    tab_options(
        table.font.size = 11,
        heading.title.font.size = 14,
        heading.subtitle.font.size = 12,
        column_labels.font.weight = "bold"
    )

```