---
title: "Understanding community assemblage using species and functional diversity across elevational gradients from the tropics to the arctic"
format:
    html:
      self-contained: true
  # pdf:
  #   documentclass: article
editor: visual
execute:
  echo: false
  error: true
  out-width: '100%'
---

```{r}
#| label: set-up
#| echo: false
#| warnings: false

library(targets)
library(tarchetypes)

```

## Introduction

Species and functional diversity are different metrics describing biodiversity and ecosystem function and they are affected by both abiotic and biotic filtering as well as the regional species pool. Species richness and functional diversity are expected to decrease towards high elevation and the arctic, due to stronger environmental filtering and fewer species in the species pool. These metrics can be used to understand how plant communities respond to different filters and how communities assembly along broad environmental gradients.

We assessed plant species diversity (i.e. species richness, evenness) and functional diversity (i.e. community weighted means and variance in leaf functional traits) along five elevational gradients in the arctic (Svalbard), boreal (Norway), , temperate (Colorado), sub-tropics (China) and tropic mountains (Peru). Specifically, we ask how plant communities respond to the environmental drivers along elevational gradients in terms of species and functional diversity? And how consistent these patterns are across a broad latitudinal gradient from the tropics to the arctic?

## Methods

We used species community composition and functional leaf trait data from all PFTC courses and RMBL in Colorado. The data represent the following ecosystems/countries: arctic (PFTC4 Svalbard), boreal (PFTC6 Norway), temperate (Colorado), sub-tropics (PFTC1 and 2 China) and tropics (PFTC3, Puna project and PFCT5 Peru). There are no chemical traits from Norway yet, and Norway was excluded from the multivariate analysis.

For some ecosystems there are several gradients, representing different management or environmental conditions. In Peru there is a gradient with more recent burning (B). In Svalbard there is a gradient with nutrient input from sea birds (N). For Norway there could be a separate gradient for grazing (G), but this is not included yet. I am unsure if we want to have these gradients or not.

The following diversity indices were calculated: Shannon diversity, Shannon evenness, species richness ( number of species) and sum of cover.

Size traits were log transformed. Trait means and higher moments were calculated using a trait imputation and bootstrapping with traitstrap (Maitner et al. 2023).

```{r}
#| label: study-sites-map
#| echo: false
#| warnings: false
#| fig-height: 12

tar_load(regions_world_map)
regions_world_map

```

\newpage

## Results

### Environmental Variables

This correlation matrix shows the relationships between bioclimatic variables across all study sites. Temperature-related variables (annual temperature, temperature range, diurnal range) show strong positive correlations with each other, while precipitation variables show more complex patterns.

```{r}
#| label: bioclim-correlation
#| echo: false
#| warnings: false
#| fig-height: 8
#| fig-width: 8

tar_load(bioclim_correlation_fig)
bioclim_correlation_fig

```

\newpage

This PCA shows the main patterns in bioclimatic variation across our study sites. The first two principal components capture the major environmental gradients that drive community assembly patterns.

```{r}
#| label: bioclim-pca
#| echo: false
#| warnings: false
#| fig-height: 6

tar_load(bioclim_pca_fig)
bioclim_pca_fig

```

\newpage

### Diversity

This figure shows diversity indices (richness, Shannon diversity, and evenness) across the latitudinal gradient. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: diversity-lat
#| echo: false
#| warnings: false
#| fig-height: 4

tar_load(diversity_lat_fig)
diversity_lat_fig

```

\newpage

This figure shows diversity indices in relation to annual temperature. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: diversity-anntemp
#| echo: false
#| warnings: false
#| fig-height: 4

tar_load(diversity_anntemp_fig)
diversity_anntemp_fig

```

\newpage

This figure shows diversity indices in relation to temperature annual range. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: diversity-temprange
#| echo: false
#| warnings: false
#| fig-height: 4

tar_load(diversity_temprange_fig)
diversity_temprange_fig

```

\newpage

### Functional leaf traits

This figure shows the multivariate trait space for a subset of traits available across all regions. The ecosystems are well separated in trait space, with clear differentiation between regions.

```{r}
#| label: trait-pca-full
#| echo: false
#| warnings: false
#| fig-height: 6

tar_load(trait_pca_full_fig)
trait_pca_full_fig

```

\newpage

This figure shows the multivariate trait space including chemical traits. Norway is not part of this figure, because we do not have the chemical traits yet. The ecosystems are well separated in trait space, with the temperate and sub-tropic regions showing some overlap.

```{r}
#| label: trait-pca
#| echo: false
#| warnings: false
#| fig-height: 6

tar_load(trait_pca_fig)
trait_pca_fig

```

\newpage

This figure shows functional leaf traits in relation to latitude. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: trait-lat
#| echo: false
#| warnings: false
#| fig-height: 8

tar_load(trait_lat_fig)
trait_lat_fig

```

\newpage

This figure shows functional leaf traits in relation to annual temperature. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: trait-anntemp
#| echo: false
#| warnings: false
#| fig-height: 8

tar_load(trait_anntemp_fig)
trait_anntemp_fig

```

\newpage

This figure shows functional leaf traits in relation to temperature annual range. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: trait-temprange
#| echo: false
#| warnings: false
#| fig-height: 8

tar_load(trait_temprange_fig)
trait_temprange_fig

```

\newpage