---
title: "Understanding community assemblage using species and functional diversity across elevational gradients from the tropics to the arctic"
format:
    html:
      self-contained: false
  # pdf:
  #   documentclass: article
editor: visual
execute:
  echo: false
  error: true
  out-width: '100%'
---

```{r}
#| label: set-up
#| echo: false
#| warnings: false

library(targets)
library(tarchetypes)

```

## Introduction

Species and functional diversity are different metrics describing biodiversity and ecosystem function and they are affected by both abiotic and biotic filtering as well as the regional species pool. Species richness and functional diversity are expected to decrease towards high elevation and the arctic, due to stronger environmental filtering and fewer species in the species pool. These metrics can be used to understand how plant communities respond to different filters and how communities assembly along broad environmental gradients.

We assessed plant species diversity (i.e. species richness, evenness) and functional diversity (i.e. community weighted means and variance in leaf functional traits) along five elevational gradients in the arctic (Svalbard), boreal (Norway), , temperate (Colorado), sub-tropics (China) and tropic mountains (Peru). Specifically, we ask how plant communities respond to the environmental drivers along elevational gradients in terms of species and functional diversity? And how consistent these patterns are across a broad latitudinal gradient from the tropics to the arctic?

## Methods

We used species community composition and functional leaf trait data from all PFTC courses and RMBL in Colorado. The data represent the following ecosystems/countries: arctic (PFTC4 Svalbard), boreal (PFTC6 Norway), temperate (Colorado), sub-tropics (PFTC1 and 2 China) and tropics (PFTC3, Puna project and PFCT5 Peru). There are no chemical traits from Norway yet, and Norway was excluded from the multivariate analysis.

For some ecosystems there are several gradients, representing different management or environmental conditions. In Peru there is a gradient with more recent burning (B). In Svalbard there is a gradient with nutrient input from sea birds (N). For Norway there could be a separate gradient for grazing (G), but this is not included yet. I am unsure if we want to have these gradients or not.

The following diversity indices were calculated: Shannon diversity, Shannon evenness, species richness ( number of species) and sum of cover.

Size traits were log transformed. Trait means and higher moments were calculated using a trait imputation and bootstrapping with traitstrap (Maitner et al. 2023).

```{r}
#| label: tab-climate-summary
#| echo: false
#| warnings: false

library(gt)
library(dplyr)

# Load CHELSA climate data
tar_load(chelsa_extracted)

# Create summary table of climate variables by region
climate_summary <- chelsa_extracted |>
  group_by(region) |>
  summarise(
    `Growing Season Length (days)` = round(mean(growing_season_length, na.rm = TRUE), 1),
    `Growing Season Temperature (°C)` = round(mean(growing_season_temperature, na.rm = TRUE), 1),
    `Growing Season Precipitation (mm)` = round(mean(growing_season_precipitation, na.rm = TRUE), 1),
    `Potential Evapotranspiration (mm)` = round(mean(potential_evapotranspiration, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  # Reorder regions from north to south
  mutate(region = factor(region, levels = c("Svalbard", "Southern Scandes", "Rocky Mountains", 
                                           "Eastern Himalaya", "Central Andes", "Drakensberg"))) |>
  arrange(region)

# Create nice table with gt
climate_summary |>
  gt() |>
  tab_header(
    title = "Climate Summary by Region",
    subtitle = "Average growing season conditions across study sites"
  ) |>
  cols_label(
    region = "Region"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = everything())
  ) |>
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  )

```

```{r}
#| label: fig-study-sites-map
#| echo: false
#| warnings: false
#| fig-height: 8

tar_load(regions_world_map)
regions_world_map

```

\newpage

## Results

### Diversity

This figure shows diversity indices (richness, Shannon diversity, and evenness) across the elevational gradient. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: fig-diversity-elev
#| echo: false
#| warnings: false
#| fig-height: 4

tar_load(diversity_fig)
diversity_fig

```

### Functional leaf traits

This figure shows the multivariate trait space for a subset of traits available across all regions. The ecosystems are well separated in trait space, with clear differentiation between regions.

```{r}
#| label: fig-trait-pca-full
#| echo: false
#| warnings: false
#| fig-height: 6

tar_load(trait_pca_full_fig)
trait_pca_full_fig

```

\newpage

This figure shows the multivariate trait space including chemical traits. Norway is not part of this figure, because we do not have the chemical traits yet. The ecosystems are well separated in trait space, with the temperate and sub-tropic regions showing some overlap.

```{r}
#| label: fig-trait-pca
#| echo: false
#| warnings: false
#| fig-height: 6

tar_load(trait_pca_fig)
trait_pca_fig

```

\newpage

This figure shows functional leaf traits in relation to latitude. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: fig-trait-elev
#| echo: false
#| warnings: false
#| fig-height: 8

tar_load(trait_elev_fig)
trait_elev_fig

```

\newpage

This figure shows functional leaf traits in relation to growing season length. The solid grey line represents the marginal prediction from mixed-effects models, while points are colored by region.

```{r}
#| label: fig-trait-gsl
#| echo: false
#| warnings: false
#| fig-height: 8

tar_load(trait_gsl_fig)
trait_gsl_fig

```

## Trait Model Results

```{r}
#| label: tab-trait-results
#| echo: false
#| warnings: false

library(gt)
library(dplyr)

tar_load(trait_results)

# Create a nice table of trait model results
trait_results_table <- trait_results |>
  filter(effect == "fixed", term != "(Intercept)") |>  # Only fixed effects, exclude intercept
  mutate(
    # Clean up trait names
    trait_clean = stringr::str_replace_all(trait_trans, "_", " ") |> 
      stringr::str_to_title(),
    # Clean up predictor names
    predictor_clean = case_when(
      term == "latitude_n" ~ "Latitude",
      term == "elevation_m" ~ "Elevation",
      term == "growing_season_length" ~ "Growing Season Length (GEE)",
      term == "gsl_1981-2010_chelsa" ~ "Growing Season Length (CHELSA)",
      term == "gst_1981-2010_chelsa" ~ "Growing Season Temperature (CHELSA)",
      term == "gsp_1981-2010_chelsa" ~ "Growing Season Precipitation (CHELSA)",
      term == "pet_penman_mean_1981-2010_chelsa" ~ "Potential Evapotranspiration (CHELSA)",
      term == "mean_temperture_warmest_quarter_bioclim" ~ "Mean Temperature Warmest Quarter (WorldClim)",
      term == "precipitation_warmest_quarter_bioclim" ~ "Precipitation Warmest Quarter (WorldClim)",
      term == "diurnal_range_bioclim" ~ "Mean Diurnal Range (WorldClim)",
      TRUE ~ term
    ),
    # Round numeric values
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 2),
    p.value = round(p.value, 3),
    # Create significance indicator
    is_significant = p.value < 0.05
  ) |>
  select(trait_clean, predictor_clean, estimate, std.error, statistic, df, p.value, is_significant) |>
  arrange(trait_clean, predictor_clean)

# Create the gt table
trait_results_table |>
  gt() |>
  tab_header(
    title = "Trait Model Results",
    subtitle = "Mixed-effects models showing relationships between functional traits and environmental predictors"
  ) |>
  cols_label(
    trait_clean = "Trait",
    predictor_clean = "Predictor",
    estimate = "Estimate",
    std.error = "SE",
    statistic = "t-statistic",
    df = "df",
    p.value = "p-value"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = is_significant == TRUE
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = c(estimate, std.error, statistic, df, p.value))
  ) |>
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    column_labels.font.weight = "bold"
  ) |>
  tab_footnote(
    footnote = "Bold rows indicate significant relationships (p < 0.05)",
    locations = cells_column_labels(columns = trait_clean)
  )

```